---
title: "Vignette 1 - Getting started"
author: "Dr Annie Visser-Quinn (a.visser-quinn@hw.ac.uk)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    theme: flatly
    highlight: breezedark
  pdf_document:
    toc: yes
    toc_depth: '3'
vignette: |
  %\VignetteIndexEntry{Vignette 1 - Getting started} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(OSMtidy)

knitr::opts_chunk$set(echo = TRUE, out.height = "75%", out.width = "75%")
```

## 1. Introduction

If you've not done so already, you can install OSMtidy using the package devtools: `devtools::install_github("avisserquinn/OSMtidy")`.

This vignette provides an example of how to prepare a shapefile for use in OSMtidy. If you don't have a shapefile, OSMtidy includes example data from the UK's Ordnance Survey. 

## 2. Example data

The example data is a shapefile of wards across the UK. The data is from the UK's Ordinance Survey Open Data (https://www.ordnancesurvey.co.uk/business-government/products/boundaryline) and is made available under the UK Government’s OGL licence; see http://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/ for details.

Access the file path of the example data with the code below.

```{r}
files <- 
  system.file("extdata", "", package = "OSMtidy") %>% 
  list.files(pattern = "district_borough_unitary_ward_region", full.names = TRUE)
files
```


## 3. Reading in a shapefile

You can read in a shapefile to R using a function from the R package sf, `st_read()`. As a dependency of OSMtidy, sf should already be installed and loaded.

If using the example data, you can use the filepath saved above. Alternatively, you can paste the filepath in as a string, e.g. `"myShapefile.shp"`.

```{r, echo = FALSE}
shp <- st_read(files[[3]]) # The third element in the vector has the extension .shp
```
```{r}
shp
```

<br>

At the top of the output it says that this is a *Simple feature collection with 7125 features and 15 fields*. Each feature represents a ward in the UK. We need to choose one location to focus on - in these vignettes we will look at a ward in the City of Edinburgh (Scotland) called Leith Walk.


## 4. Tidying the shapefile

In the next code chunk, we reduce the data down to the location of interest using tidyverse notation.

```{r}
shp <- 
  shp %>%
  filter(str_detect(FILE_NAME, "CITY_OF_EDINBURGH")) %>% 
  filter(str_detect(NAME, "Leith Walk")) %>%
  select(geometry)
shp
```

<br>

Because the shapefile was imported to R as a spatial feature, using sf, we can plot it in ggplot using `geom_sf()`.

```{r, out.width="35%"}
shp %>% 
  ggplot() + 
  geom_sf() + 
  theme_void()
```

You can simplify the jagged outlines of the ward using `st_simplify()`. The final step is to transform the projection to EPSG:4326 (https://epsg.io/4326) using `st_transform()`.

```{r}
shp <- 
  shp %>% 
  st_simplify(dTolerance = 25) %>% 
  st_transform(4326)
shp
```

```{r, out.width="35%"}
shp %>% 
  ggplot() + 
  geom_sf() + 
  theme_void()
```


## 5. Export the shapefile

The shapefile is now ready for use in OSMtidy. Shapefiles can be exported using the sf function `st_write()`. 

```{r}
shp %>% 
  st_write("exampleEdinburgh.shp", 
           delete_dsn = TRUE, 
           delete_layer = TRUE, 
           quiet = TRUE)
```

<br>

Now you’re ready to run OSMtidy! See Vignette 2 for a general walkthrough of the OSMtidy workflow.
