---
title: "Vignette 2 - OSMtidy workflow"
author: "Dr Annie Visser-Quinn (a.visser-quinn@hw.ac.uk)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    theme: flatly
    highlight: breezedark
  pdf_document:
    toc: yes
    toc_depth: '3'
vignette: |
  %\VignetteIndexEntry{Vignette 2 - OSMtidy workflow} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(OSMtidy)

knitr::opts_chunk$set(echo = TRUE, out.height = "75%", out.width = "75%")
```

This vignette describes the OSMtidy workflow. The workflow consists of six steps which are intended to be simple and easy to follow:

1.	**Input** A shapefile outlining the location
2.	**Extract** Spatial data – inside the shapefiles ‘bounding box’ – is extracted from [OpenStreetMaps](https://www.openstreetmap.org/) servers via the R package [osmdata](https://cran.r-project.org/web/packages/osmdata/index.html)
3.	**Cut** The extracted data is ‘cookie cutter’-ed to the shapefile extent
4.	**Wrangle** The data is transformed into a suitable format for filtering
5.	**Filter** The physical objects are filtered and renamed to follow a simple naming convention
6.	**Tidy** Collates the outputs to form a streamlined database of physical objects

This vignette also introduces to helper functions which can be applied at all of these six steps: `dataSummary()` and `dataExport()`.

To get started, you'll need:

- An up to date install of the OSMtidy package. You can install or update OSMtidy using the package devtools: `devtools::install_github("avisserquinn/OSMtidy")`
- A shapefile (a single polygon) outline of the location you want to extract OSM data for


## 1. Input - Using dataShapefile()

Using the function `dataShapefile()` we can import the shapefile for which data is to be extracted. There are two input arguments:

- *filename* The filename or filepath of the shapefile
- *crs* Coordinate projection of the input data (optional). This function will automatically convert the shapefile to EPSG:4326 (https://epsg.io/4326)

```{r}
shp <- dataShapefile(filename = "exampleEdinburgh.shp", crs = 4326) # or dataShapefile("exampleEdinburgh.shp")
shp
```


### 1.1 dataSummary()

At each step, you can print a summary of the OSMtidy outputs using the function `dataSummary()`.

```{r, out.width="35%"}
dataSummary(shp)
```


### 1.2 dataExport()

You can also export any OSMtidy output using the function `dataExport()`. The export file names have the following convention:

- Location name
- Step number (e.g. 1)
- Step name (e.g. dataShapefile)
- A timestamp of when the output was exported

```{r}
dataExport(shp, "exampleEdinburgh")
```

## 2. Extract - Using dataExtract()

The OpenStreetMap data is extracted, via the R package osmdata and the overpass server, using the function `dataExtract()`. 


### 2.1 Features

This data is extracted by feature, and a vector of the names of features consisting of physical objects can be generated via `data("features")`. A vector of all the vailable features in osmdata can be accessed via the function `osmdata::available_features`.

```{r}
data("features")
features <- features[c(2,18,19,22,24)]; features # For this example we'll select a subset of 5 features

osmdata::available_features # A vector of all 209 available features
```

### 2.2 dataExtract()

The function `dataExtract()` has three input arguments: 

- *dataShapefile* The shapefile output from step 1
- *timeout* The time in seconds before the query to the overpass server will time out; see *Details* below
  - Default 300 seconds
- *memsize* The memory size for the overpass server; see *Details* below
  - Default 1073741824 Bytes
  
<details><summary><span style = "color: #2196F3;">**Details from the R package osmdata**</span></summary>
  
> **timeout** It may be necessary to increase this value for large queries, because the server may time out before all data are delivered.
> **memsize** The default memory size for the 'overpass' server in bytes; may need to be increased in order to handle large queries.>
> See https://wiki.openstreetmap.org/wiki/Overpass_API#Resource_management_options_.28osm-script.29 for explanation of timeout and memsize (or maxsize in overpass terms). Note in particular the comment that queries with arbitrarily large memsize are likely to be rejected.
</details>

<br>

**Note** this function may take some time to run. Timestamps and progress are printed while the function is running. It is recommended that you execute the function once to avoid flooding the overpass server. The example below extracts 5 of the 47 features. 

```{r, cache = TRUE}
dlExtract <- dataExtract(dataShapefile = shp, features = features)

dataSummary(dlExtract)

dataExport(dlExtract, "exampleEdinburgh")
```

## 3. Cut - Using dataCut()

In step 2 the data was extracted as a "bounding box" (a rectangle). In step 3, the data is cut to the shapefile using the function `dataCut()`. 

The function `dataCut` has two input arguments: 

- **dataExtracted** Output from step 2
- **dataShapefile** Output from step 1

Timestamps and progress are printed when the function is running. 

```{r, cache = TRUE}
dlCut <- dataCut(dataExtracted = dlExtract, dataShapefile = shp)

dataSummary(dlCut)

dataExport(dlCut, "exampleEdinburgh")
```


## 4. Wrangle - Using dataWrangle()

Using the function `dataWrangle` we can tidy up (or wrangle) the data before filtering. 

There is one input argument: 

- **dataCut** Output from step 3

Timestamps and progress are printed when the function is running. 

```{r, cache = TRUE}
dlWrangle <- dataWrangle(dataCut = dlCut)

dataSummary(dlWrangle)

dataExport(dlWrangle, "exampleEdinburgh")
```


## 5. Filter - Using dataFilter()

### 5.1 Filters

The main function of OSMtidy is `dataFilter()`. Here, the data is filtered based on rules set out in an Excel spreadsheet. Default filters, generated as part of the Water Resilient Cities project, can be accessed using the `data()` function: `data("filters")`.

```{r}
data("filters")
filters
```

To see the spreadsheet the default filters are based on, or to access a template to create your own filters, generate the filepaths using the code below.

```{r}
system.file("extdata", "filters.xlsx", package = "OSMtidy")
system.file("extdata", "filtersTemplate.xlsx", package = "OSMtidy")
```

See vignette 3 for further details. 

### 5.2 filterOverview()

A filter overview can be generated using the function `filterOverview()`. The input can be either the filepath as a string or as an object in R. Examples of both are provided below.

<details><summary><span style = "color: #2196F3;">**filterOverview() with a filepath**</span></summary>

```{r}
filepath <- system.file("extdata", "filters.xlsx", package = "OSMtidyPackage")
filepath
filterOverview(filepath)
```

</details>

<details><summary><span style = "color: #2196F3;">**filterOverview() with an R object**</span></summary>

```{r}
data("filters")
filters
filterOverview(filters)
```

</details>

<br>


### 5.3 Application

There are three input arguments to `dataFilter()`:

- **dataWrangle** Output from step 4
- **filters** The filepath of the filters as a string or as an object in R (see `filterOverview()` previously)
- **rows** Specify the rows in the filters object to apply. Intended for troubleshooting and adjusting filters
  - Default to NULL, i.e. all filters (rows)
  
Timestamps and progress are printed when the function is running. 

Depending on the location size, number of filters and computer performance, filters can take anything from a couple of minutes (the example ward) to multiple hours to run (City of London and Boroughs). 

```{r message=FALSE, warning=FALSE, cache=TRUE}
dlFilter <- dataFilter(dataWrangle = dlWrangle, filters = filters)
dataSummary(dlFilter)
dataExport(dlFilter, "exampleEdinburgh")
```


## 6. Tidy - Using dataTidy()

Note that multiple outputs from `dataWrangle()` and `dataFilter()` were spreadsheets (.xlsx extension). These spreadsheets may be manually adjusted; this is covered in the third vignette.

The function `dataTidy()` generates a single tidied output based on any combination (R object and/or spreadsheet) of this filtered, validated, unfiltered and no detail data. 

There is one input argument to `dataTidy()`:

- **datalist** A list of R objects and/or spreadsheet. 

In this vignette, the input is the list of outputs from steps 4 and 5. In vignettes 3 and 4 a number of alternative approaches are described. 

**The tidied geotagged dataset is saved in .RDS, and .csv for use in a range of applications. To export as a shapefile it is necessary to split the geotagged dataset by geometry type first. Still need to update how the final output works and runs **

```{r}
dlTidy <-
  dataTidy(dataList = 
             list(dlWrangle$noDetail,
                  dlFilter$unfiltered,
                  dlFilter$filtered,
                  dlFilter$validate))
dataSummary(dlTidy)
dataExport(dlTidy, "exampleEdinburgh")
```
